\textbf{NBA Party}

### 2017-07-11 cardinal_analytx_presentation/nightlife_and_sports.tex
Restructuring. Still a long way to go.

% What is our solution?

\begin{frame}   
  \frametitle{Why is this problem hard?}
  \begin{block}{Partying is a latent variable.}
    Instead, we proxy for the possibility of athletes partying by looking at 
    play following games in party cities. Our question becomes...
  \end{block}
  \vspace{12pt}\begin{block}{Do teams exhibit a decline in performance the day following a game in a city with active nightlife?}     I.e. we are concerned with ``hangover effects''.
  \end{block}
  \vspace{12pt}\begin{block}{Part of the question is determining a measure for nightlife. }
    We colloquially call cities with high levels of nightlife ``party cities''.   \end{block}
\end{frame}

\begin{frame}   
  \frametitle{Who cares?}
  \begin{block}{\$400 billion spent on sports betting per year in the US.\footnote{NBA Commissioner Adam Silver, 2014.}}

@@

\begin{frame}
  \frametitle{Previous research - NBA player shooting average}
  \begin{block}{Empirical evidence: players party to the detriment of the game.}
  NBA Players under-perform proportion of shots made on Sundays.\footnote{Chase, ('15).}
  \end{block}
    \begin{itemize}       
    \item 1,092 of 2,035 NBA players who took $\geq 1$ shot on Sundays under-perform 
      relative to their own season-long average; $p=0.001$.
      \item Effect size is not considered.
      \item Density of support is questionable.\footnote{What if we miss the only shot we took on Sunday, is that relevant?}     \end{itemize}
\end{frame}

\begin{frame}   \frametitle{Prev. research - likelihood of meeting the spread in NBA}
  \begin{block}{Empirical evidence: away teams under-perform on Sundays.}
  Teams under-perform bookmakers expectations on Sunday afternoons when on the road.\footnote{Ezekowitz, ('14).}
    \begin{itemize}       
    \item Only 129 out of 284 games did the away team meet the spread, 
      using 2006-14 data; $p=0.1089$.
    \item Not quite significant at the 90\% confidence level.     
    \item If we bet \$100 on each Sunday game, we'd only profit \$1,310.\footnote{$(284 - 129) \times 100 + 129 \times (-110) = 1,310$.} 
    \end{itemize}
  \end{block}
  \end{frame}

@@

    A team is expected to meet the spread $\frac{1}{2}$ the time.\footnote{Dubner and Levitt, ('06). Losers pay 10\% fee on top of wagers; a balanced book guarantees 5\% gain for the house-bookmaker.}
  \end{block}
  \vspace{12pt}\begin{block}{Bookmaker money-lines (MLB)}     The odds that a team will beat their opponent.

@@

@@

\begin{frame}   \frametitle{Sports: NBA and MLB}
  \begin{block}{We perform the analysis for both NBA and MLB}     
    If we observe a team traveling to a city with active nightlife:
    \vspace{12pt}
    \begin{itemize}       \item For NBA teams playing another game in the next 24 hours, are they less likely to meet the points-spread?
      \item For MLB teams, after conditioning on the odds of winning, are they less likely to beat their opponent?     \end{itemize}
    \vspace{12pt}\begin{block}{If we yield a causal model, we may realize a betting schema.}     
\end{block}
  \end{block} 
\end{frame}

@@

\begin{frame}   \frametitle{Next day opponent: NBA}
  \centering \includegraphics[scale=0.25]{../writing/next_day_opponent} \end{frame}

\begin{frame}   %\frametitle{Further motivation for conditioning on the spread}
  \begin{block}{With some correlation between next-day opponent, this further motivates conditioning
  on bookmakers' expectations.}
  \begin{itemize}     \item If bookkeepers aware of hangover effects, our variable of interest may not appear significant, in spite of there being a true effect.

    \item If bookkeepers systematically mis-value teams which are more likely to be encountered after visiting a city with active nightlife, we could see a spurious effect.

      (Given size of betting market, this is unlikely)
    \end{itemize}   \end{block}
\end{frame}

\begin{frame}   \frametitle{Recall: partying is a latent variable}
  \begin{block}{We can't observe when players go for a night on the town.}
    Instead, we develop two approaches to proxy for partying.

\begin{frame}   \frametitle{Discrete Indicator}
  \begin{block}{Anecdotally, LA and NY appear in news articles frequently.}     \vspace{12pt}This motivates a binary indicator for party when either of these two cities have been visited within 24-hours of another game.
\vspace{12pt}
    \[
      \texttt{party} \coloneqq \mathbbm 1 \left \{\substack{\textrm{last game played in LA or NY} \\ \textrm{at most 24-hours ago}}\right \}
    \]   \end{block} \end{frame}

@@

\begin{frame}   \frametitle{Defense in the NBA}
  \begin{adjustwidth}{-.75cm}{}
  \tiny{  % mpa, tpa, tps
\begin{tabular}{@{\extracolsep{5pt}}lccc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]   & \multicolumn{3}{c}{\textit{Dependent variable:}} \\  \cline{2-4}  \\[-1.8ex] & Team Points Admitted & Team Points Admitted & Team Points Scored \\  \\[-1.8ex] & (1) & (2) & (3)\\  \hline \\[-1.8ex]   Discrete party indicator & 2.817$^{***}$ &  &  \\    & (0.712) &  &  \\    & & & \\   Continuous party measure &  & 1.942$^{***}$ & 0.377 \\    &  & (0.248) & (0.246) \\    & & & \\   Lag change in possessions &  & 0.022$^{***}$ & 0.035$^{***}$ \\    &  & (0.007) & (0.007) \\    & & & \\   Logged travel distance &  & $-$0.411$^{***}$ & 0.035 \\    &  & (0.074) & (0.073) \\    & & & \\   Number hours since last game &  & 0.012$^{***}$ & 0.020$^{***}$ \\    &  & (0.003) & (0.003) \\    & & & \\   Constant & $-$0.046 & 102.824$^{***}$ & 99.540$^{***}$ \\    & (0.091) & (0.483) & (0.478) \\    & & & \\  \hline \\[-1.8ex]  Observations & 15,139 & 44,689 & 44,689 \\  R$^{2}$ & 0.001 & 0.002 & 0.002 \\  Adjusted R$^{2}$ & 0.001 & 0.002 & 0.002 \\  Residual Std. Error & 11.137 (df = 15137) & 12.117 (df = 44684) & 12.015 (df = 44684) \\  F Statistic & 15.670$^{***}$ (df = 1; 15137) & 27.968$^{***}$ (df = 4; 44684) & 24.325$^{***}$ (df = 4; 44684) \\  \hline  \hline \\[-1.8ex]  \textit{Note:}  & \multicolumn{3}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\  \end{tabular}  }
\end{adjustwidth} 
\end{frame}

>>>>

\begin{frame}   
  \frametitle{Why is this important?}
  \begin{block}{\$400 billion spent on sports betting per year in the US.\footnote{NBA Commissioner Adam Silver, 2014.}}

@@

\begin{frame}   
  \frametitle{Why is this problem hard?}
  \begin{block}{Partying is a latent variable.}
    Instead, we proxy for the possibility of athletes partying by looking at 
    play following games in party cities.
  \end{block}
  \vspace{12pt}\begin{block}{Part of the question is determining a measure for nightlife. }
    We colloquially call cities with high levels of nightlife ``party cities''.   \end{block}
\end{frame}

\begin{frame}   \frametitle{What's the specific question, exactly?}
  \begin{block}{Do teams/athletes perform worse the day after playing in a party-oriented city?\footnote{  We will define ``perform worse'' relative to bookmaker expectations, which are in part driven by
  bettors' expectations.}}
\vspace{12pt}     I.e. we are concerned with ``hangover effects''.
  \end{block}
\end{frame}

@@

    A team is expected to meet the spread $\frac{1}{2}$ the time.\footnote{Dubner and Levitt, ('06). Losers pay 10\% fee on top of wagers; a balanced book guarantees 5\% gain for the house-bookmaker.}
  \end{block}
  \vspace{12pt}
  \begin{block}{Bookmaker money-lines (MLB)}     
    The odds that a team will beat their opponent.

@@

\begin{frame}   \frametitle{Anecdotal evidence: players do party}
  \begin{block}{Open attitude toward partying}     Matt Barnes recently complained, ``there's no nightlife in Utah'' when asked
    about playing the Jazz in the 2017 playoffs.\footnote{Cestone and Seiter ('17).}   \end{block}

  \vspace{12pt}
  \begin{block}{Blaming the party scene for poor performance}
    J.R. Smith's career-to-date success rate of shots taken was 42\%, 
    but on Sunday afternoon games he only managed 34\% in 2012-13.\footnote{Concepcion ('15).}

    \vspace{12pt}
    After being traded from NY to Cavaliers, performance increased.
    When asked why: ``There’s nothing you expect (in Cleveland) but basketball. 
    There’s nothing, there’s no going out, there’s no late nights.''\footnote{Price, Levy ('15).}
  \end{block} \end{frame}

\begin{frame}
  \frametitle{Previous research - NBA player shooting average}
  \begin{block}{Empirical evidence: players party to the detriment of the game.}
  NBA Players under-perform proportion of shots made on Sundays.\footnote{Chase, ('15).}
  \end{block}
    \begin{itemize}       
    \item 1,092 of 2,035 NBA players who took $\geq 1$ shot on Sundays       under-perform 
      relative to their own career-long average;\footnote{Analysis limited to all players who entered the league after 1974.} $p=0.001$.
    \end{itemize}
    \begin{block}{Problems}       \begin{itemize}           \item Effect size is not considered.
          \item Density of support is questionable.\footnote{What if we miss the only shot we took on Sunday, is that relevant?}            \end{itemize}     \end{block}
\end{frame}

\begin{frame}   \frametitle{Prev. research - likelihood of meeting the spread in NBA}
  \begin{block}{Empirical evidence: away teams under-perform on Sundays.}
  Teams under-perform bookmakers expectations on Sunday afternoons when on the road.\footnote{Ezekowitz, ('14).}
    \begin{itemize}       
    \item Only 129 out of 284 games did the away team meet the spread, 
      using 2006-14 data; $p=0.1089$.
    \end{itemize}
    \vspace{12pt}
    \begin{block}{Only consider Sunday games $\leadsto$ underwhelming results:}       \begin{itemize}
          \item Not quite significant at the 90\% confidence level.     
          \item If we bet \$100 on each Sunday game, we'd only profit \$1,310 over 8 years.\footnote{$(284 - 129) \times 100 + 129 \times (-110) = 1,310$.} 
      \end{itemize}     \end{block}
  \end{block}
  \end{frame}

@@

\begin{frame}   \frametitle{Returning to our question within the NBA and MLB}
  \begin{block}{We perform the analysis for both NBA and MLB}     
    If we observe a team traveling to a city with active nightlife:
    \vspace{12pt}
    \begin{itemize}       \item For NBA teams playing another game in the next 24 hours, are they less likely to meet the points-spread?
      \item For MLB teams, after conditioning on the odds of winning, are they less likely to beat their next opponent?     \end{itemize}
    \vspace{12pt}\begin{block}{If we yield a causal model, we may realize a betting schema.}
\end{block}
  \end{block} 
\end{frame}

\begin{frame}   \frametitle{Overzealous assumption: next day opponent uncorrelated}
  \centering \includegraphics[scale=0.25]{../writing/next_day_opponent} \end{frame}

\begin{frame}
  \frametitle{Weakening our assumptions}
  \begin{block}{With some correlation between next-day opponent, this further motivates using bookmaker expectations as the response.}

    \vspace{12pt}
    Helps to account for systematically running into teams the day after visitation to 
    a party oriented city who are particularly (not)-skilled.
  \end{block}
\end{frame}

\begin{frame}   \frametitle{Public scrutiny makes data collection difficult}

  \begin{block}{Whether parties take place is a latent variable}     Due to intense public scrutiny, there is a reluctance to share nightlife experiences
    of players on social media.

  \end{block} 
  \vspace{12pt}
  \begin{block}{We can't observe when players go for a night on the town.}

    Instead, we develop two approaches to proxy for partying.

\begin{frame}   \frametitle{Discrete indicator}
  \begin{block}{Anecdotally, LA and NY appear in news articles frequently.}     \vspace{12pt}This motivates a binary indicator for party when either of these two cities have been visited within 24-hours of another game.
\vspace{12pt}
    \[
      \texttt{party} \coloneqq \mathbbm 1 \left \{\substack{\textrm{last game played in LA or NY} \\ \textrm{at most 24-hours ago}}\right \}
    \]   \end{block} \end{frame}

  % \begin{itemize}     \item If bookkeepers aware of hangover effects, our variable of interest may not appear significant, in spite of there being a true effect.

  %   \item If bookkeepers systematically mis-value teams which are more likely to be encountered after visiting a city with active nightlife, we could see a spurious effect.

  %     (Given size of betting market, this is unlikely)
  %   \end{itemize}

@@

@@

\begin{frame}   \frametitle{Defense in the NBA}
  \begin{adjustwidth}{-.75cm}{}
  \tiny{  % mpa1, tpa, tps
\begin{tabular}{@{\extracolsep{5pt}}lccc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]   & \multicolumn{3}{c}{\textit{Dependent variable:}} \\  \cline{2-4}  \\[-1.8ex] & Team Points Admitted & Team Points Admitted & Team Points Scored \\  \\[-1.8ex] & (1) & (2) & (3)\\  \hline \\[-1.8ex]   Discrete party indicator & 2.649$^{***}$ &  &  \\    & (0.718) &  &  \\    & & & \\   Continuous party measure &  & 1.833$^{***}$ & 0.447$^{*}$ \\    &  & (0.246) & (0.245) \\    & & & \\   Lag change in possessions & 1.058$^{*}$ & 0.022$^{***}$ & 0.035$^{***}$ \\    & (0.615) & (0.007) & (0.007) \\    & & & \\   Logged travel distance & $-$0.045 & $-$0.028 & $-$0.213$^{***}$ \\    & (0.038) & (0.074) & (0.074) \\    & & & \\   Number hours since last game & 0.009$^{**}$ & 0.015$^{***}$ & 0.018$^{***}$ \\    & (0.004) & (0.003) & (0.003) \\    & & & \\   Home team effect & $-$2.997$^{***}$ & $-$3.412$^{***}$ & 2.211$^{***}$ \\    & (0.218) & (0.116) & (0.116) \\    & & & \\   Constant & $-$3.429 & 101.941$^{***}$ & 100.112$^{***}$ \\    & (2.926) & (0.479) & (0.477) \\    & & & \\  \hline \\[-1.8ex]  Observations & 14,843 & 44,689 & 44,689 \\  R$^{2}$ & 0.018 & 0.021 & 0.010 \\  Adjusted R$^{2}$ & 0.017 & 0.021 & 0.010 \\  Residual Std. Error & 11.070 (df = 14831) & 12.002 (df = 44683) & 11.967 (df = 44683) \\  F Statistic & 24.021$^{***}$ (df = 11; 14831) & 196.086$^{***}$ (df = 5; 44683) & 92.809$^{***}$ (df = 5; 44683) \\  \hline  \hline \\[-1.8ex]  \textit{Note:}  & \multicolumn{3}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\  \end{tabular}    }
\end{adjustwidth} 
\end{frame}
